Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2020-02-18 paper-bedpca.tex
loadings ordered
minor corrections on text

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite{yang2010common,loh2015contrasting}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix. These includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality. Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these. 
In the following, we outline these pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic association within these LD regions.  
Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
This is also an issue when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing various statistics. We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values. 
We use these new functions to analyse the UK Biobank data, and show that they are both very fast and easy to use. We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure. 
Interestingly, subsetting the UK Biobank data enables to get more PCs that capture population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects or family structure, or when the goal is to restrict the data to individuals of homogeneous ancestry.

@@

\subsection{Efficient implementation of PCA for genotype data}

When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$). Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).

When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing values of each variant.
Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}. 
This algorithm is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector, which we implement for PLINK bed files and parallelise.

@@

Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance of $x$.
We use these distances for many applications in this paper. 
When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
If $x$ are PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters, as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}. 
We reexport function \texttt{covRob} in R package bigutilsr for convenience.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
Then, we compute $K$ PC scores and loadings.
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window of 50 variants as radius by default).

Finally, to choose the threshold on the previously described statistics above which variants are considered outliers, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their respective KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster. 
Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detects as outlier any point that is next to a very dense cluster.
We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics; see e.g.\ figure \ref{fig:outlier-pd}).

@@

To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).

@@

We provide and use a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with second-degree relationship or more, we apply KING-relatedness cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are made available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
X-axis represent positions of variants in the data, and y-axis the value of loadings.
Points are hex-binned to make plotting of such large number of points manageable.
\label{fig:UKBB-loadings40}}
\end{figure}

As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique. It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).

@@

To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability''). 
In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.55\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics (S)}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
A few points with higher values for this statistic S appear as outliers in PC17-20.}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

As for restricting to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
Here, we visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table \ref{tab:homogeneous}). 
In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.

@@

We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute K=20 PCs. Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings; 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with \texttt{bed\_tcrossprodSelf}, one of the new functions of R package bigsnpr; and 3/ the OADP projection (section \ref{proj}). 
When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
In contrast, there is no visible shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this bias (Figure \ref{fig:proj1000G-2}).
We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than with simple projection, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP. Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
\label{proj-related}

@@

When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}). Overall, this shrinkage decreases with an increased sample size (Table \ref{tab:shrinkage}).

@@

We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unknown self-reported ancestry. 
We further removed all pairs of related individuals reported by the UKBB (i.e.\ both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (\texttt{bed\_projectSelfPCA}). Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB. Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

\subsection{Imputation of missing values}

>>>>

Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts. For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}. PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.

However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix. These steps includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness. Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these. 
In the following, we outline these pitfalls and explain when they are relevant.
First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,zou2010quantification,abdellaoui2013population,prive2017efficient}.
Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic associations within these LD regions \cite[]{zou2010quantification}.  
Second, another issue may arise when projecting a new study dataset to the PCA space computed from a reference dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}. This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
This same issue also arises when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs. This potential issue has been demonstrated for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.

For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing various statistics. We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values. 
We use these new functions to analyse the UK Biobank data, and show that these functions are both very fast and easy to use. We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure. 
Interestingly, subsetting the UK Biobank data enables to get more PCs that capture population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to get unbiased projections instead.
Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects or distant family structure, or when the goal is to restrict the data to individuals of homogeneous ancestry.

@@

\subsection{Efficient implementation of PCA for genotype data \label{randomSVD}}

When there is no missing value, we compute the truncated Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$). Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).

When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing genotypes.
Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}. 
This algorithm, implemented in R package RSpectra, requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector. We implement such multiplication in parallel from a PLINK bed file.

@@

Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance.
We use these distances for many applications in this paper. 
When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
If $x$ represents PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust}.
We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
Then, we compute $K$ PC scores and loadings (Section \ref{randomSVD}).
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances of these PC loadings (Section \ref{maha}).  
To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window with a radius of 50 variants by default).

Finally, to choose the threshold on the previously described statistics above which variants are considered outliers, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts for both skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their respective KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). Intuitively, an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster. 
Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ as statistic to detect individual outliers in PCA. Note the square root, as it otherwise detects as outlier any point that is next to a very dense cluster.
We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics; see e.g.\ figure \ref{fig:outlier-pd}).

@@

To project a target genotype dataset to the PCA space from a reference dataset (e.g.\ the 1000 genomes data), we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
To project individuals from the same dataset as the individuals used for computing PCA, we provide function \texttt{bed\_projectSelfPCA}. 
Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).

@@

We provide and use a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with second-degree relationship or more, we apply KING-relatedness cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are made available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
To easily download this data, we provide function \texttt{download\_1000G} in R package bigsnpr.

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
Column indices of variants in the data, ordered by chromosome and physical position, are represented on the x-axis, and the value of loadings are represented on the y-axis.
Points are hex-binned.
\label{fig:UKBB-loadings40}}
\end{figure}

As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique. It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the most computationally demanding step when computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).

@@

To detect a few outlier samples, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
All but one outlier are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability''). 
In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
We recall that, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, the probability of detecting one outlier or more using 6SD is only of 0.0001.

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.55\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics (S).}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
A few points with higher values for this statistic S appear as outliers in PC17-20.}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

As for restricting to homogeneous samples, we compare the use of R package aberrant, which was used to report the homogeneous ``White British'' subset in UKBB, to the use of the robust Mahalanobis distance we propose here.
We visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table \ref{tab:homogeneous}). 
In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.

@@

We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute K=20 PCs. Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings; 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with \texttt{bed\_tcrossprodSelf}, one of the new functions of R package bigsnpr; and 3/ the OADP projection (section \ref{proj}). 
When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
In contrast, there is no visible shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this bias (Figure \ref{fig:proj1000G-2}).
We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than with simple projection, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP. Projections of related individuals using OADP suffers from some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
\label{proj-related}

@@

When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}). Overall, this shrinkage for simple projection decreases with an increased sample size (Table \ref{tab:shrinkage}).

@@

We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unknown self-reported ancestry. 
We further removed all pairs of related individuals reported by the UKBB (i.e.\ both individuals in each pair). 
This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (function \texttt{bed\_projectSelfPCA}). Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in the UKBB. Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).

@@

\subsection{PCA \& missing value imputation}
